Twenty years of dynamic occupancy models: a review of applications and a look towards the future

### 2024-09-15 Manuscript.qmd
Addressed introduction, methods, and some result comments

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing can be applied in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy a difficult quantity to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with most wildlife field data. Generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017]. Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses for tasks including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites exist in either an occupied or unoccupied state. In the first time step, occupancy state at each site is determined by the probability of initial occupancy. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation and local extinction. In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability describes whether or not the species is observed during a survey; under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur.

@@

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this allows for considerably greater flexibility in data inputs.

@@

DOMs make a number of key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need for authors to consider how certain their detections truly are. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably met, model extensions designed to account for false positive error should be considered.

II. **Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion around the bias which is introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites. A more pertinent approach for most studies, however, is careful consideration of ecologically relevant seasons corresponding to an appropriate definition of occupancy.

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.

Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To capture any additiona relevant articles which did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles which did not meet inclusion criteria were replaced from within their own strata.

As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles which fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.

For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach. To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims. The possible categories were *Observing trends*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing relationships*, where authors explore specific predefined hypotheses of relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Predicting temporally*, where authors predict site occupancy under future conditions, *Predicting spatially*, where authors predict site occupancy to unsurveyed locations, and *Developing methods*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.

@@

We were particularly interested in how authors navigated the modelling process, from covariate selection through model evaluation. To this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models. Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]). Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter; and *simple precursors*, where selection was preceded by another simpler model implementation. Finally, for each model we documented whether goodness-of-fit was tested and reported, as well as whether model performance was assessed by validation with either in or out-of-sample data. For the full spreadsheet of data collected and further details on categorisation, see supplementary material.

@@

A total of 92 articles were included for this review. Based on the acceptance rates within each strata and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).

@@

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how some this variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and an exceptional diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranging from under one month to forty years, with a median of 6 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

DOMs have been used to study a variety of different species, though the majority of studies have been conducted on data for vertebrate taxa. They have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 uses them to model the spread of chronic wasting disease. The latter's application to disease dynamics is not unique, and DOMs have been touted as a valuable tool for modelling disease dynamics. @mores2020 and @padilla-torres2013 have each used DOMs to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats. The vast majority of applications have been fit for terrestrial species, though there have been a limited number of studies on aquatic systems, including @fisher2014's application to invasive salmon, @falke2012's on Great Plains stream fishes, and @pendleton2022's on whale occupancy dynamics.

While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. While they did not fit multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.

>>>>

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. An understanding of where a species exists, which factors lead to its presence, and how its distribution may be changing is applicable in many ways. Descriptions of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions. Robust knowledge of the occupancy patterns of a species also helps us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios.

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models [@gutiérrez-arellano2024] commonly used to study meta-population dynamics, and species distribution models [SDMs, @franklin2010] widely used to explore species occurrence at larger scales. However, several factors can make occupancy difficult to estimate. For instance, simple presence/absence data of species observations can be biased when species are detected imperfectly, as is the case with much wildlife field data — generally speaking, it is impossible to determine whether a location is truly unoccupied or whether the species occurs but was not detected by the observer [@lahoz-monfort2014; @gu2004]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species' occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions often occur during biological invasions and climate change driven range shifts, both conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] . Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes which drive occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@kéry2013; @briscoe2021].

@@

The DOM structure encompasses two processes: the ecological process of site occupancy dynamics describing the presence or absence of a species at a site at any point in time, and the observational process of detection describing whether a species is observed at a site where it is present (@fig-modelform). Sites exist in either an occupied (z = 1) or unoccupied state (z = 0). In the first time step, occupancy state at each site is determined by the probability of initial occupancy ψ~1~. In following time steps, occupancy state can change according to a Markovian process where occupancy is predicated on the site's state in the prior time-step and the probabilities of colonisation γ and local extinction ε (@eq-occupancy).

$$
\begin{align*}
z_1 &\sim Bernoulli(\psi_1)\\\\
z_t &\sim Bernoulli(\psi)\\
&\psi = z_{t-1}(1-\epsilon)+(1-z_{t-1})\gamma
\end{align*}
$$ {#eq-occupancy}

In the observation component of the DOM, the model explicitly accounts for imperfect detection. At occupied sites, the detection probability ρ describes whether the species is observed during a survey (y = 1) or not (y = 0). Under the DOM's original parameterisation, it is assumed that false-positive detections at unoccupied sites never occur (@eq-detection).

$$
y_{t,j} \sim Bernouilli(z_t\rho)
$$ {#eq-detection}

@@

To disentangle the ecological and observational processes the DOM requires a hierarchical sampling design as depicted in @fig-surveys. Under this design, observations at a site occur during distinct, time-bound seasons within which it is assumed site occupancy does not change (that is, sites are closed to changes in occupancy). In each season multiple observations are conducted, permitting estimation of the probability of detection conditional on occupancy. Most frequently these repeat observations are collected by revisiting the site on separate occasions, although they can also be attained by alternative means: examples include conducting surveys at multiple locations within a site, using multiple observers during a survey, or recording the time elapsed until a detection is recorded [@guillera-arroita2017a; @guillera-arroita2017b]. It is important to note that it is not necessary for the same number of observations to occur in each year or for each site, and that the DOM accommodates for missing data; this enables flexibility in data inputs. In this review we use the term "parameter" to refer to the quantities being estimate: in most cases, these are initial occupancy ψ~1~, colonisation γ, extinction ε, and detection ρ (see @fig-modelform, @eq-occupancy, and @eq-detection), though some alternative model forms may directly estimate occupancy ψ as a parameter.

@@

DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need to consider how certain detections truly are for any given study system. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably be met, model extensions designed to account for false positive error should be considered [@royle2006; @miller2011; @miller2015].

II. **Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias which is introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and @valente2017 discuss the difficultly in distinguishing between temporary emigration and local extinction. Model extensions to relax the closure assumption have been developed, including @kendall2013's approach using staggered arrival and departure periods between sites. A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@kéry2021].

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species depends not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.

Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of the most common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

@@

To assess how DOMs have been applied in the years since their introduction we gathered a representative sample of articles fitting them to field ecological data. A pool of candidate articles was generated using two queries on Web of Science. The first of these included all articles from 2004-2023 which cite @mackenzie2003. To capture any additional relevant articles that did not directly cite @mackenzie2003, a second query was generated searching articles in the same time-span matching the terms "*dynamic occupancy model\**", "*multi-season occupancy model\**", or "*occupancy dynamic\**"; articles including each of "*occupancy*", "*colonization*", "*extinction/persistence*", and "*detection*"; and articles with the term "*occupancy*" located near "*dynamic*" in the title, key terms, or abstract. As we were interested in how DOMs use has changed through time, we divided all articles across four-year-long strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2024. From each of these strata we randomly selected 20 articles for inclusion in the review. Articles that did not meet inclusion criteria were replaced from within their own strata.

As our review focuses only on applications of the dynamic multi-season occupancy model of @mackenzie2003 and its extensions, we included articles that fit a model meeting the following criteria: i) uses non-simulated, field-collected, presence-absence data; ii) relies on data from multiple sites which can exist in at least two states, including an occupied and unoccupied state; iii) has multiple seasons, between which sites may change states conditional on the prior season's occupancy state and transition probabilities such as colonisation and extinction; and iv) contains at least one parameter describing the detection process.

For each article we recorded key details on authorship, research objectives, study taxa and system, survey methods, and modelling approach. To examine the reasons why authors used DOMs, we allocated each article to one or more category of objective based on the study's stated aims. The possible categories were *Estimating trends*, where authors express interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; *Testing relationships*, where authors explore specific predefined hypotheses of relationships between covariates and model parameters; *Identifying drivers*, where authors attempt to find which covariates influence model parameters; *Predicting temporally*, where authors predict site occupancy under future conditions, *Predicting spatially*, where authors predict site occupancy to unsurveyed locations, and *Developing methods*, where authors introduce, test, or demonstrate aspects of dynamic occupancy models.

@@

We were particularly interested in how authors navigated the modelling process, from covariate selection through model evaluation. To this end, we recorded all covariates considered in each study regardless of whether they were or were not included in final models. Key traits of each covariate were recorded including their general category, whether they were directly observed or remotely sensed, whether they were static or varied between seasons, and how they were presented in the model (as a linear term, a polynomial term, or as part of an interaction with another covariate [@james2021]). Model selection procedures were also sorted into non-exclusive categories including *A priori*, where only one model was considered; *candidate suite*, where a predefined set of models was considered; *sequential,* where covariates were selected parameter-by-parameter (e.g., fitting detection first, followed by initial occupancy and so on); and *simple precursors*, where selection was preceded by another simpler model implementation. Finally, for each model we documented whether goodness-of-fit was tested and reported, and whether model performance was assessed by validation with either in or out-of-sample data. For the full spreadsheet of data collected and further details on categorisation, see supplementary material.

@@

A total of 92 articles were included for this review. Based on the acceptance rates within each stratum and quantity of unprocessed articles, an estimated 496 of the 1152 unreviewed articles in our sample would have met inclusion criteria (@fig-coverage).

@@

Articles included in our review demonstrate considerable diversity in scope, scale, and objectives. A selection of key attributes of these studies is presented in @fig-StudyDetails, and we provide notable examples of DOMs throughout this section to highlight how the noted variation appears in practice. Study systems included in our sample are globally distributed; while a majority use data collected in the United States of America, representatives are included from all geographic realms and a broad diversity of ecosystems. A particularly notable aspect of these study areas is their variation in size, which ranges from the hyper-local to the continental scale. The smallest study locality included in our sample studied insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@basset2023; @clement2019].

The temporal scale of studies shows similar variability. The duration between first and last survey ranged from under one month to forty years, with a median of 6 years. More meaningful for model interpretation, however, is the duration of the primary occasion, as this is presumed to be the temporal scale at which changes in occupancy are observed. Most applications describe the primary occasion as a year, although some studies divide years into multiple primary occasions to describe seasonal variation in occupancy. In the most extreme cases the primary occasion may be as brief as a week, capturing much finer scale changes in occupancy. These shortened-season occupancy models are most common with camera trapping or acoustic monitoring data, which can be arbitrarily divided into primary and secondary sampling occasions. This is seen in @kleiven2020's study of rodent and mustelid interactions using camera trap data, where the primary occasions are less than one week in length. On the other end of the spectrum, some studies modelled primary occasions which were decades apart and represented generational changes in occupancy. This is illustrated in @couturier2023's study on long-term otter recovery in France, which used data from two primary occasions: one in 2003 and one in 2012.

The DOM has been used to study a variety of different species, though the majority of studies have focussed on vertebrate taxa (@fig-StudyDetails B). The DOM has been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there are exceptions — @belinchón2017 fits a DOM to lichen data, and @cook2022 model the spread of chronic wasting disease. The latter's application to disease dynamics is not unique, and the DOM has been touted as a valuable tool for such applications. @mores2020 and @padilla-torres2013 have used the DOM to model mosquito dynamics as a human disease vector, and @bailey2014 discusses the concept of applying the DOM to the spread of chytrid fungus in amphibian habitats. The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic systems, including invasive salmon [@fisher2014], Great Plains stream fishes [@falke2012], and whales [@pendleton2022].

While most articles in our sample fit a model to a single species, 41 fit models to multiple species either as independent models or explicitly multi-species implementations. The multi-species models use various extensions of the conventional DOM, including hierarchical models which fit hundreds of species in a single implementation with species-level effects[@dorazio2010; @hendershot2020], as well as explicit interaction models which estimate conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019]. Rather than using multi-species models, several other authors fit large numbers of independent models to different species [@peach2019; @otto2012]. Working with large numbers of taxa does raise additional challenges, as the level of complexity which can realistically be considered for each individual taxon is likely to be reduced for the sake of practicality.
